---
title: "Coursera Data Analysis Project "
date: "October 30, 2015"
output:
  html_document:
    theme: cerulean
---

<!-- Enter the code required to load your data in the space below. The data will be loaded but the line of code won't show up in your write up (echo=FALSE) in order to save space-->
```{r echo=FALSE}
load(url("http://bit.ly/dasi_gss_data"))
```

<!-- In the remainder of the document, add R code chunks as needed -->

Based on the General Social Survey, Cumulative File 1972-2012 [1], the analysis below will answer the question whether there is a relationship between the highest level of education and the number of children the person had. It is an interesting to see if higher levels of education have an influence (among many other factors) on the size of the family.

The analyis uses the data about the people surveyed in 2000-2012 who had completed their degree and had children (excluding people who had a degree and didn't have children).


### Data:

The data were collected from 28 General Social Surveys from the year 1972 to 2012, conducted at the National Opinion Research Center at the University of Chicago. In 2013, this data were brought together as the General Social Survey,Cumulative File 1972-2012. This was an observational study and the data were collected based on the information provided by the respondents, sampled from all noninstitutionalized English and Spanish speaking persons 18 years of age or older, living in the United States.

The results can be generalized to this subset of population, because the population was randomly sampled (multistage sampling method). The potential sources of bias are non-response cases (working parents are hard to reach during workdays), and, possibly, parents in a process of obtaining a higher degree who reported the last degree they obtained. These data cannot be used to establish a causal link between a degree and the number of children, because causation can only be established from an experiment with random assignment, and also because there are other factors affecting the number of children a person has (e.g., health condition or lifestyle choice). 

For the analysis, the explanatory variable is the person's highest level of education (categorical with five levels) and the response variable is the number of children (discrete numerical).

As seen from the contigency table below, there are five degree groups: "Lt High School, "High School",  "Junior College",  "Bachelor", and  "Graduate", and the  number of children within each group is from one to eight.

```{r}
millenium<-subset(gss, gss$year>=2000 & gss$childs>=1)
table(millenium$childs,millenium$degree)
```

*Note:* "LT High School" stands for "Less than High School" [1, p.153].


### Exploratory data analysis:

The summary statistics and side-by-side box plots for each degree group follow.

```{r}
by(millenium$childs, millenium$degree, summary)
boxplot(millenium$childs ~ millenium$degree, xlab="Highest degree", ylab="Number of children")
```

All box-and-whiskers plots have a natural lower boundary at 1, because it is not possible to have less than zero kids, and, additionally, only respondents with children (i.e. one child or more) are considered. 

The plots indicate that the distributions within groups are right-skewed, with majority of the data located near a lower boundary with noticable outliers on the left.  

The plots suggest that there might be a correlation between these two variables, i.e. people with higher degrees tend to have fewer children than people with lower degrees.

### Inference:

To define the null and alternative hypotheses for the test, the question is formulated as follows: is there a difference between the average number of children for people with different degrees? 

Since more than two means are considered, analysis of variance (ANOVA) method is used.

* $\sf{H_{0}}$:The mean number of children is the same across all degrees;
* $\sf{H_{A}}$: At least one pair of means is different from each other.

ANOVA conditions check:

*  Independence:
    + Between groups - the groups are independent (non-paired);
    + Within each group - the respondents were randomly sampled [1, Appendix A], and sample size of each group is less than 10% of the population;
  
* Approximate normality - as noticed earlier by box-and-whiskers plots visual check, the distributions within groups are right-skewed. The sample size is relatively large, so such skew can be explained by the presense of the natural boundary on the right. Analysis of the normal probability plots [2] also indicates that there is a right tail in all the distributions with noticable outliers beyond three standard deviations away from the mean.

```{r eval=TRUE, echo=FALSE}
lths_set<-subset(millenium$childs, millenium$degree=="Lt High School")
hs_set<-subset(millenium$childs, millenium$degree=="High School")
jk_set<-subset(millenium$childs, millenium$degree=="Junior College")
b_set<-subset(millenium$childs, millenium$degree=="Bachelor")
g_set<-subset(millenium$childs, millenium$degree=="Graduate")
qqnorm(lths_set, main = "Lt High School")
qqline(lths_set)
qqnorm(hs_set, main = "High School")
qqline(hs_set)
qqnorm(jk_set, main = "Junior College")
qqline(jk_set)
qqnorm(b_set, main = "Bachelor")
qqline(b_set)
qqnorm(g_set, main = "Graduate")
qqline(g_set)
```

* Equal variance - roughly equal variablity between groups, see below.

```{r}
by(millenium$childs, millenium$degree, sd)
```


```{r eval=TRUE, echo=FALSE}
source("http://bit.ly/dasi_inference")
```

#### ANOVA:

```{r}
inference(millenium$childs, millenium$degree, est="mean", type="ht", method="theoretical", alternative="greater", siglevel = 0.005, eda_plot=TRUE, inf_plot=TRUE, sum_stats=TRUE)
```

*Note:* The alternative parameter is set to "greater", as if ommited, the following message is displayed in the console:

```{r eval=FALSE, echo=TRUE}
Warning message:
Use alternative = 'greater' for ANOVA.
```

The Bonferroni correction of the significance level $\alpha$ is used for multiple comparisons. This new significance level $\alpha^*$ is determined as described below.

Number of comparisons to made between groups is K
$$
K=\frac{k(k-1)}{2},
$$

where $k$ is the number of groups, and it equals $5$ in this case.

The modified significance level for multiple comparisons $\alpha^*$ is

$$
\alpha^*=\frac{\alpha}{K}.
$$

```{r}
k<-5
K<-(k*(k-1))/2
alpha<-0.05
alpha_star<-alpha/K
alpha_star
```


### Conclusion:

The p-value is almost zero, so we reject the null hypothesis in favour of the alternative hypothesis and conclude that there is a difference in the average number of children for the people with different degrees, and it is not due to sampling variation  or chance. 

Examining pairwise tests results, we can make a conclusion that there is a statistically significance difference between the mean number of children for the people who completed Lt High School and people with all other degree levels, even considering the stringent significance level $\alpha^*=0.00$, since p-values in the first column are all zeros.
The picture for the people with High School degree is similar. However, the p-value for comparison of number of children of parents with High School and Junior College $0.0053$, which is very close but still greater than the $\alpha^*$,  and we can interpret it as there is no statistically significant difference int the mean number of children between these two groups. The p-values are again zeros for High School - Bachelor and High School - Graduate pairs, meaning statistically significant difference in the average number of kids. 
In the last two columns of the pairwise comparisons table for Junior College-Bachelor, Junior College-Graduate, and Bachelor-Graduate pairs the p-values are considerably larger than $\alpha^*$, and it tells us that there is no statistically significant difference in the mean number of children between these groups.

Summarizing all the above, it can be concluded that, statistically, people with the lower (up to high school) degrees have on average more children than people with higher degrees (college or university). It also can be noticed that the above analysis shows that there is no difference in the average number of children in the families where parents obtained different degrees after completing the hight school. Thus, it is possible to say that based on the GSS Cumulative Survey for people interviewed between 2000 and 2012, there is a relationship between the level of education (i.e. high shool or lower vs. secondary/post-secondary) and the number of children the person had.

As the level of educations is only a one factor among many others that have an effect the number of children, the extension of this research may include the analysis of the relationship between the number of children and, for example, the family total income, or the religious believes, or the average age of both parents.

### References

[1] [Smith, Tom W., Michael Hout, and Peter V. Marsden. General Social Survey, 1972-2012 (Cumulative File). ICPSR34802-v1. Storrs, CT: Roper Center for Public Opinion Research, University of Connecticut /Ann Arbor, MI: Inter-university Consortium for Political and Social Research (distributors), 2013-09-11. doi:10.3886/ICPSR34802.v1] (http://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/34802/version/1)

[2] [http://www.pmean.com/09/NormalPlot.html] (http://www.pmean.com/09/NormalPlot.html)

### Appendix:

The first 100 cases of the data set are represented below.

```{r eval=TRUE, echo=FALSE}
mydata<-data.frame(millenium$caseid[1:100], millenium$degree[1:100], millenium$childs[1:100])
names(mydata)<-c("Case ID", "Degree", "# of children")
na.omit(mydata)
```


